---
title: "Predictive Maintenance: Capstone Project Report"
subtitle: "HarvardX PH125.9x - Data Science: Capstone"
author: "Seyed M. Seyedtorabi"
date: "2024-Feb-16"
output: pdf_document
---

\newpage

```{r libraries and functions, include=FALSE}
# installing the libraries
if(!require(tidyverse   )) install.packages("tidyverse"  , repos = "http://cran.us.r-project.org")
if(!require(caret       )) install.packages("caret"      , repos = "http://cran.us.r-project.org")
if(!require(corrplot    )) install.packages("corrplot"   , repos = "http://cran.us.r-project.org")
if(!require(GGally      )) install.packages("GGally"     , repos = "http://cran.us.r-project.org")
if(!require(ROCR        )) install.packages("ROCR"       , repos = "http://cran.us.r-project.org")
if(!require(klaR        )) install.packages("klaR"       , repos = "http://cran.us.r-project.org")
if(!require(xgboost     )) install.packages("xgboost"    , repos = "http://cran.us.r-project.org")
if(!require(pROC        )) install.packages("pROC"       , repos = "http://cran.us.r-project.org")
if(!require(smotefamily )) install.packages("smotefamily", repos = "http://cran.us.r-project.org")
if(!require(knitr       )) install.packages("knitr"      , repos = "http://cran.us.r-project.org")

library(knitr) 
library(corrplot)
library(tidyverse)
library(caret)
library(GGally)
library(ROCR)
library(pROC)
library(smotefamily)
library(UBL)
library(MASS)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
################################################################################
############### Defining functions to be used in script#########################
################################################################################

# function for one_hot encoding of categorical feature (type in our case)
one_hot_encode <- function(df_data, categorical_col){
  # Ensure the specified column is categorical
  if (!is.factor(df_data[[categorical_col]])) {
    stop("Specified column is not categorical.")
  }
  
  # Get unique categories in the categorical column
  categories <- levels(df_data[[categorical_col]])
  
  # Create binary columns for each category
  for (category in categories) {
    binary_col <- ifelse(df_data[[categorical_col]] == category, 1, 0)
    col_name <- paste(categorical_col, category, sep = "_")
    df_data[[col_name]] <- binary_col
  }
  
  # Return the encoded dataframe
  df_data
}

################################################################################
train_model <- function(df_training, target,
                        features, method, control){
  # declare the formula (y ~ x1 + x2 + etc.)
  formula <- as.formula(
    paste(
      target, "~", paste(features, collapse = " + ")))
  
  # train the model
  model <- train(formula,
                 data = df_training,
                 method = method,
                 trControl = control)
  model
}

################################################################################
get_confusion_matrix <- function(model, df_test, target) {
  # make confusion matrix
  cm <- confusionMatrix(predict(model, df_test),
                        df_test[[target]])
  cm
}
################################################################################
get_ROC_performance <- function(model, df_test, target) {
  # get probabilities
  predicted_probs <- predict(model, newdata = df_test, type = "prob")[, "No Failure"]
  pred <- prediction(predicted_probs, df_test[[target]])
  # get tpr and fpr to plot ROC
  perf <- performance(pred, "tpr", "fpr")
  perf
}
################################################################################
get_tpr_at_x_fpr <- function(model, df_test, target, x=0.1) {

  perf <- get_ROC_performance(model=model, df_test=df_test, target=target)
  
  # Extract TPR and FPR values
  fpr <- perf@x.values[[1]]
  tpr <- perf@y.values[[1]]
  
  # Interpolate to find TPR at the specified FPR value
  approx_tpr <- approx(fpr, tpr, xout = x)$y
  approx_tpr
}
################################################################################
plot_ROC_curves <- function(trained_models, model_names, df_test, target, title) {
  
  # Create an empty plot
  plot(0, 0, type = "n", xlim = c(0, 1), ylim = c(0, 1), 
       xlab = "False Positive Rate", ylab = "True Positive Rate", 
       main = title)
  
  grid(col = "grey", lty = "dotted")
  # Plot ROC curves for each model
  for (i in seq_along(trained_models)) {
    perf <- get_ROC_performance(trained_models[[i]], df_test, target)
    auc <- round(as.numeric(perf@y.values[[1]]), 3)  # Extract AUC value
    
    # Extract TPR and FPR values
    fpr <- perf@x.values[[1]]
    tpr <- perf@y.values[[1]]
    
    # Plot ROC curve
    lines(fpr, tpr, col = i, lwd = 2)
  }
  
  # Add legend
  legend("bottomright", legend = model_names, 
         col = seq_along(trained_models), lwd = 2, cex = 0.8)
}
################################################################################
plot_ROC_curves_cross_comparison <- function(trained_models_1, 
                                             trained_models_2,
                                             set_1_description,
                                             set_2_description,
                                             method_names,
                                             model_names, 
                                             df_test, target,
                                             df_test_2 = data.frame()) {

  
  # create a dictionary using setNames()
  method_model_dictionary <- setNames(model_names, method_names)
  
  # Create an empty plot
  plot(0, 0, type = "n", xlim = c(0, 1), ylim = c(0, 1), 
       xlab = "False Positive Rate", ylab = "True Positive Rate", 
       main = "ROC curves comparison")
  
  grid(col = "grey", lty = "dotted")
  legend_names <- character()
  
  # use same test set if another one is not given
  if (nrow(df_test_2)==0){df_test_2 <- df_test}
  
  color_index <- 1
  # plot ROC curves for the methods which are in "method_names" argument
  for (i in seq(1, length(trained_models_1))) {
    
    method_name_1 <- trained_models_1[[i]][["method"]]
    
    for (ii in seq(1, length(trained_models_2))) {
      
      method_name_2 <- trained_models_2[[ii]][["method"]]
      
      if (method_name_1==method_name_2 & method_name_2 %in% method_names){
        
        perf_1 <- get_ROC_performance(trained_models_1[[i]], 
                                      df_test, 
                                      target)
        
        perf_2 <- get_ROC_performance(trained_models_2[[ii]], 
                                      df_test_2, 
                                      target)
        
        # extract TPR and FPR values
        fpr_1 <- perf_1@x.values[[1]]
        tpr_1 <- perf_1@y.values[[1]]
        fpr_2 <- perf_2@x.values[[1]]
        tpr_2 <- perf_2@y.values[[1]]
        
        # plot ROC curve
        lines(fpr_1, tpr_1, col = color_index, lwd = 2, lty=1)
        lines(fpr_2, tpr_2, col = color_index, lwd = 2, lty=2)
        
        # make the right legends
        legend_names <- c(legend_names, 
                          paste(set_1_description, 
                            method_model_dictionary[[method_name_1]]))
        
        legend_names <- c(legend_names, 
                          paste(set_2_description, 
                            method_model_dictionary[[method_name_1]]))
        
        color_index <- color_index + 1
        }
      }
    }

  # Add legend
  legend("bottomright", legend = legend_names, 
         col = rep(seq_along(model_names), each = 2), 
         lwd = 2, lty = c(1, 2), cex = 0.8)
}

################################################################################
# Function to perform SMOTE on failure type, over sample based on type frequency
smote_for_failure_type <- function(train_data, target_col, positive_class, 
                                   subclass_type_col, features_cols,
                                   apprx_subclass_pop) {
  
  # Extract features and target variables
  train_features <- train_data[, features_cols]
  train_failure_type <- train_data[[subclass_type_col]]
  
  # Calculate the frequency of each Failure type
  failure_type_freq <- table(train_failure_type)
  
  # most common item in the failure type column (No Failure)
  majority_class <- names(which.max(failure_type_freq))
  
  # Find minority Failure types, exclude the most common one (No Failure)
  #failure_types <- names(failure_type_freq[failure_type_freq != majority_class])
  failure_types <- setdiff(names(failure_type_freq), c(majority_class))
  
  # Perform SMOTE for each minority Failure type
  df_smote <- list()
  i <- 1
  for (failure_type in failure_types) {
    # Subset the data for the current Failure type
    subset_data <- train_data[train_failure_type == failure_type |
                                train_failure_type == majority_class, ]
    
    # find the correct dupsize based on the desired subclass population
    dupsize <- apprx_subclass_pop/failure_type_freq[[failure_type]]
    
    # Perform SMOTE on the subset
    smote_subset <- SMOTE(subset_data[, features_cols], 
                          target = subset_data[[target_col]],
                          dup_size = dupsize)
    
    # take the dataset with over-sampled minority and add a failure type column
    df_smote_subclss <- smote_subset$data
    df_smote_subclss[[subclass_type_col]] <- ifelse(
      df_smote_subclss$class == positive_class, 
      failure_type, majority_class)
    
    # Rename the class column to match the original target column name
    df_smote_subclss <- df_smote_subclss %>% 
      rename(!!target_col := class)
    
    # Combine SMOTE results for each Failure type
    df_smote[[i]] <- df_smote_subclss
    i <- i+1
  }
  
  # Combine SMOTE results for all minority Failure types
  df_smote <- do.call(rbind, df_smote)
  
  # remove duplicated rows due to majority class
  df_smote <- distinct(df_smote)
  
  # make both target columns factors again
  df_smote[[target_col]] <- as.factor(df_smote[[target_col]])
  df_smote[[subclass_type_col]] <- as.factor(df_smote[[subclass_type_col]])
  
  # return the final dataframe
  df_smote
}

################################################################################
# Function to perform DBSMOTE on failure type, over sample based on type frequency
DBSMOTE_for_failure_type <- function(train_data, target_col, positive_class, 
                                     subclass_type_col, features_cols,
                                     apprx_subclass_pop) {
  
  # Extract features and target variables
  train_features <- train_data[, features_cols]
  train_failure_type <- train_data[[subclass_type_col]]
  
  # Calculate the frequency of each Failure type
  failure_type_freq <- table(train_failure_type)
  
  # most common item in the failure type column (No Failure)
  majority_class <- names(which.max(failure_type_freq))
  
  # Find minority Failure types, exclude the most common one (No Failure)
  #failure_types <- names(failure_type_freq[failure_type_freq != majority_class])
  failure_types <- setdiff(names(failure_type_freq), c(majority_class))
  
  # Perform SMOTE for each minority Failure type
  df_smote <- list()
  i <- 1
  for (failure_type in failure_types) {
    # Subset the data for the current Failure type
    subset_data <- train_data[train_failure_type == failure_type |
                                train_failure_type == majority_class, ]
    
    # find the correct dupsize based on the desired subclass population
    dupsize <- apprx_subclass_pop/failure_type_freq[[failure_type]]
    
    # Perform SMOTE on the subset
    smote_subset <- DBSMOTE(subset_data[, features_cols], 
                            target = subset_data[[target_col]],
                            dupSize = dupsize)
                            #K = 5,
                            #method="type2")
    
    # take the dataset with over-sampled minority and add a failure type column
    df_smote_subclss <- smote_subset$data
    df_smote_subclss[[subclass_type_col]] <- ifelse(
      df_smote_subclss$class == positive_class, 
      failure_type, majority_class)
    
    # Rename the class column to match the original target column name
    df_smote_subclss <- df_smote_subclss %>% 
      rename(!!target_col := class)
    
    # Combine SMOTE results for each Failure type
    df_smote[[i]] <- df_smote_subclss
    i <- i+1
  }
  
  # Combine SMOTE results for all minority Failure types
  df_smote <- do.call(rbind, df_smote)
  
  # remove duplicated rows due to majority class
  df_smote <- distinct(df_smote)
  
  # make both target columns factors again
  df_smote[[target_col]] <- as.factor(df_smote[[target_col]])
  df_smote[[subclass_type_col]] <- as.factor(df_smote[[subclass_type_col]])
  
  # return the final dataframe
  df_smote
}
################################################################################
# Function to perform ENN on failure type, in order to reduce noise in the data
apply_ENN <- function(df, target, features, 
                      k = 5, dist = "Euclidean") {
  
  df <- df[c(features, target)]
  
  # Prepare the formula
  formula <- as.formula(paste(target, "~", 
                              paste(features, collapse = " + ")))
  
  # Apply ENN
  res_enn <- ENNClassif(formula, df, k = k, dist = dist, Cl = "all")
  
  # Remove instances identified by ENN
  data_filtered <- df[-res_enn[[2]], ]
  
  data_filtered
}
################################################################################
# converting multiclass classification to binary classification
conver_to_binary_confusion_matrix <- function(multiclass_confusion_matrix){
  
  # Extract true negatives (TN)
  TN <- multiclass_confusion_matrix["No Failure", "No Failure"]
  
  # Calculate false positives (FP)
  FP <- sum(multiclass_confusion_matrix[, "No Failure"][-2])
  
  # Calculate false negatives (FN)
  FN <-  sum(multiclass_confusion_matrix["No Failure", -2 ])
  
  # Calculate true positives (TP)
  TP <- sum(diag(multiclass_confusion_matrix)) - TN
  
  # Construct the binary confusion matrix
  binary_confusion_matrix <- matrix(c(TP, FP, FN, TN), nrow = 2, byrow = TRUE,
                                    dimnames = list(c("Predicted Failure", 
                                                      "Predicted No Failure"),
                                                    c("Failure", "No Failure")))
  binary_confusion_matrix
}
################################################################################
# calculates f1 score after converting multiclass classification to binary
get_f1_score<- function(binary_confusion_matrix){
  
  precision <- binary_confusion_matrix[1,1]/(binary_confusion_matrix[1,1] + binary_confusion_matrix[1,2])
  recall <- binary_confusion_matrix[1,1]/(binary_confusion_matrix[1,1] + binary_confusion_matrix[2,1])
  F1 <- 2 * ( (precision*recall) / (precision+recall) )
  F1
}
################################################################################
# function hard voting ensemble 
predict_ensemble <- function(models, ensemble_methods, new_data) {
  
  predictions <- sapply(ensemble_methods, function(method) {
    predict(models[[which(methods==method)]], new_data)
  })
  
  # Combine predictions using majority voting
  majority_vote <- apply(predictions, 1, function(row) {
    names(which.max(table(row)))
  })
  majority_vote
}
```


# I. Introduction
This document is the generated report for the "Chose Your Own Project" capstone of the "HarvardX PH125.9x: Data Science: Capstone" course provided by HarvardX. In the intoducion section of this document, we will go over the motivations and goals behind the project, the chosen data set, problem definition, and the adapted strategy to approach the problem. In later sections, we will go over the exploratory data analysis (EDA), different modelling strategies, model performance comparisons and final results. The document will end with concluding remarks and suggestions of possible improvements for a potential future work.

## I.I. Motivation
As the previous "Movielens" capstone project was essentially a regression problem, it was desired to have a classification problem for the final capstone project. Working on a **predictive maintenance** project presented a relevant and intellectually stimulating prospect for this capstone.
Predictive maintenance is a growing field of interest in data-driven engineering. The goal of predictive maintenance is to monitor the health state of an asset (e.g. a machine, a component thereof, or even a part of the infrastructure) through analysis of the data collected by sensors which monitor the asset of interest. For instance, if vibration levels of a component is under continuous monitoring by a sensor which sends the data to a database, and the machine learning models detect a change point or a trend in these vibration levels, the asset owner can send inspection personnel to investigate the alert, and possibly carry a preemptive maintenance action to avoid total failure of the asset. In this way, the resources of the inspection personnel is used more efficiently, because the periodic and/or reactive maintenance is replaced by the more targeted approach of **predictive maintenance**. In this project, we will work on a data set to classify the state of machines and predict which failure mode they have based on the provided data.

## I.II. The Data set
The Data set chosen for this project is the **Machine Predictive Maintenance Classification** data set available on Kaggle website [1]. We can import the data set and have a first look after renaming the columns to convenient names:
```{r import data and show head, echo=TRUE}
# load data set
df_data <- read.csv("./data/predictive_maintenance.csv")

# rename columns to more convenient names
df_data <- df_data %>% 
  rename("product_id" = Product.ID,
         "type" = Type,
         "air_t" = Air.temperature..K.,
         "process_t" = Process.temperature..K.,
         "rpm" = Rotational.speed..rpm.,
         "torque" = Torque..Nm.,
         "wear" = Tool.wear..min.,
         "failure_numeric" = Target,
         "failure_type" = Failure.Type)

# show the head
knitr::kable(head(df_data), row.names = FALSE)
```
After a quick check we found out that the first two columns are the unique identifiers for every row. Therefore, they are not relevant for any modelling considerations.

According to the data set description [1], the **type** column describes the type of the machine and contains possible 3 values of L, M, and H, which stand for low, medium, and high quality, respectively. the **air_t** and **process_t** columns contain respectively the air temperatures and the machine processing temperatures in Kelvin. The **rpm** column stores the angular speed of the machine in revolutions per minute (rpm), while the **torque** columns contains the angular force in Nm. The **wear** column indicates the tool wear in minutes. The last 2 columns are the labels. The column **failure_numeric** contains binary values, 1 for machine failure and 0 for no failure, followed by the **failure_type** columns which indicates the mode of failure. Here, we can see the failure modes:
```{r failure modes, echo=TRUE}
unique(df_data$failure_type)
```
We can also check if the data set is balanced or not by defining and looking at the distribution in the **failure** column:
```{r unbalanced dataset, echo=TRUE}
# crate the failure label column form 0 and 1 ==> no failure and failure
df_data <- df_data %>%
  mutate(failure = ifelse(failure_numeric==1, "Failure", "No Failure"))

table(df_data$failure)
```
The data set contains 9661 observations without failure, and only 339 failure observations. This shows that the data set is quite unbalanced and a classification algorithm may be affected by this.

## I.III. The Objective
The objective in this project is to use different data analysis, data transformation, feature engineering, and machine learning methods to classify and predict the failure of the machines. Although there are 6 distinct categories in the **failure_type** column, one of them is essentially the negative case ("No Failure"), while the other categories are the positive cases. In other words, this classification problem is inherently a binary problem from this point of view. Additionally, it is possible to imagine a real-world scenario where the machine owner is more interested in the health status of the machine rather than the specific failure mode.
It is important to mention that in our modelling efforts, we may use the multi-class classification approaches, but if the model predicts e.g. **Heat Dissipation Failure** while the real case is **Overstrain Failure**, we will still count it as a true positive. 
Since the data set is very unbalanced, we will not use **accuracy** as the metric to evaluate the model performance. Instead, we will use **F1-score** for that purpose. As a reminder, the F1-score is calculated through the following formula:

\[
F1 = 2 \times\frac{recall \times precision}{recall+precision}
\]

Precision and recall are defined by the following formulas:

\[
Precision = \frac{TP}{TP+FP}
\]

\[
Recall = \frac{TP}{TP+FN}
\]

Additionally, we will use comparisons of receiver operating characteristic (ROC) curves to visualize the model performance. One more metric that will be calculated throughout some of the sections in this report is the true positive rate (TPR) at 10% false positive rate (FPR). This metric can helpful, because in a real-world scenario, the machine owner may require the data scientist to produce a model that does not have more than 10% false alarm rate, while maintaining a high true alarm rate. Please note that TPR is simply another name for recall, and FPR is defined as:
\[
FPR = \frac{FP}{TN+FP}
\]

As mentioned before, the objective in this project make the binary failure/no failure classification. We will try to achieve an F1-score of **0.90**. We will also use the ROC curves and the TPR at 10% FPR metric as helpful means that guide our modelling efforts.

\newpage


# II. Analysis
## II.I. EDA
Prior to any modelling step, it is crucial to perform some level of EDA and visualization in order to be aware of the relationships and correlations that exist in our data. This will be the starting point in our journey to create a classification model. As the first step, let us create a detailed figure including multiple plots, so that we will see the big picture more clearly:

```{r pair plot, echo=TRUE, fig.height=7, fig.width=10}
# select columns for pair plot
df_pair <- df_data %>% 
  dplyr::select(-any_of(c("UDI", "failure", "failure_numeric",
                   "product_id", "failure_type")))

# create and show the pair plot
pair_plot <- ggpairs(df_pair,
                     mapping=aes(color=df_data$failure),
                     lower = list(continuous = wrap("points", alpha = 0.2), 
                                  combo = "box_no_facet"),
                     diag = list(continuous = wrap("densityDiag", alpha = 0.2))) +
  labs(title ="Feature pair-wise relations grid plot") + 
  theme(axis.text.x = element_text(angle = 90,))

pair_plot
```
```{r removing df_pair, include=FALSE}
# remove df_pair from environment as it's no longer needed
rm(df_pair)
```
This is a very dense figure, but one can already draw very useful insights from it. The first thing to notice is that there are many more blue points compared to red points in the subplots. This means the data set is extremely unbalanced and has a lot more "No Failure" cases than "Failure" ones. From the bar plot at the top left, one can see that L type machines have the most failures, followed by M and H type machines. The box plots on the left and top of the figure show us some rpm, torque, and wear have substantially different distributions for failed and not failed machines, regardless of the machine type. This is also visible in the density plots. The air and process temperatures also seem to be higher in case of failed machines. The figure also shows the correlation coefficients of each feature pair in both failure and no failure cases. In some pairs such as **air_t** vs **wear**, the difference in correlations for the 2 classes seem to be very high. However, one needs to take into consideration that the correlation values for failure cases are based on much smaller number of samples, and different failure types might actually have different correlation values. It is also clear from the figure that the relationship between torque and rpm is inverse. This is expected, because in machines higher torque always comes with a lower rotational speed assuming a constant power output.

Next, let us have a look at the distribution of the failure modes:
```{r failure mode bar plot, echo=TRUE, fig.height=5, fig.width=10}
# see number of different types of failure
df_data %>% 
  filter(failure_numeric==1) %>%
  group_by(failure_type) %>% 
  summarize(n_failure = n()) %>% 
  mutate(failure_type=reorder(failure_type,n_failure)) %>%
  ggplot(aes(x=failure_type, y=n_failure)) + 
  geom_col(color = "steelblue", fill="blue3") + 
  labs(x = "Failure type",
       y = "Number of failures",
       title ="Failure type vs Number of failures") + 
  coord_flip()
```

There are 2 issues one can observe in this plot: \newline
1) "No Failure" appears in the plot despite we filtered them out.\newline
2) "Random Failures" are not in the plot despite the exist in the data set as we discussed in the introduction section.

Therefore, we need to perform some data cleaning. We change the **failure_numeric** column to 0 and **failure** column to "No Failure" for all instances with "No Failure" in **failure_type** columns. We also change the failure to to "No Failure" for all instances with 0 in the **failure_numeric** column:

```{r data cleaning 1, echo=TRUE}

# reassign correct failure and failure_numeric to the problematic rows
df_data <- df_data %>% 
  mutate(failure_numeric = ifelse(failure_type=="No Failure", 0, 1),
         failure = ifelse(failure_type=="No Failure", "No Failure", failure))

# reassign correct failure_type to the random failure rows
df_data <- df_data %>% 
  mutate(failure_type = ifelse(failure=="No Failure", 
                               "No Failure", 
                               failure_type))
```
```{r convert to factor, include=FALSE}
# convert categorical features into factors instead of strings
df_data$type <- as.factor(df_data$type)
df_data$failure <- as.factor(df_data$failure)
df_data$failure_type <- as.factor(df_data$failure_type)
```
Next, let us visualize the correlations in a better way. This time we also want to include the machine type. However, the machine type in out data set is categorical and we cannot calculate correlation for non-numeric variables. To overcome this, we do **one-hot encoding** of the type data, i.e. we encode the type data in a numeric way. After this encoding, we will have 3 new columns in our data set. Here, we see how these columns are related to the **type** column:
```{r one hot encoding, echo=TRUE}
# one-hot encoding the categorical type column using the self-defined function
df_data <- one_hot_encode(df_data = df_data, categorical_col = "type")

# show encoding
tail(df_data %>% dplyr::select(type, type_L, type_M, type_H))
```
Now that we have numerically encoded the machine types, we can make create a comprehensive correlation plot. Please note that the correlation coefficients are calculated using the Spearman method.
```{r column rearrangement, include=FALSE}
# re-arrange columns to have the target columns at the end 
all_cols <- names(df_data)
target_cols <- c("failure", "failure_type", "failure_numeric")
other_cols <- setdiff(all_cols, target_cols)
df_data <- df_data[, c(other_cols, target_cols)]
```
```{r correlation plot, echo=TRUE, fig.height=6, fig.width=10}
# select columns for correlations plot
df_corr <- df_data %>% 
  dplyr::select(-any_of(c("UDI", "type", "failure",
                   "product_id", "failure_type")))

# plot the correlations using spearman method
ggcorr(df_corr, label = TRUE, label_round=3, 
       method=c("pairwise", "spearman"),
       low = "brown2", 
       mid = "lightgrey", 
       high = "steelblue")
```
```{r delete the corr plot, include=FALSE}
# remove df_corr from environment as it's no longer needed
rm(df_corr)
```
In this plot, we can see that higher torque and wear are correlated with failure. Also, rpm is inversely correlated. There is some variation in correlation coefficients between failure and different machine types, however, the coefficients are too small to conclude anything.

## II.II. Splitting the Data
Now that we have an initial understanding of our data, it is time to prepare the data set for training models. The first step in that direction would be to split our data into a test set and a training set:
```{r seed, include=FALSE}
set.seed(1)
```
```{r split data, echo=TRUE}
# Split the data set into training and test sets
test_indices <- createDataPartition(df_data$failure, times=1, p=0.2, list=FALSE)
test_set     <- df_data[test_indices, ]
train_set    <- df_data[-test_indices,]
```
We use a 80/20 split for our project. This is to make sure we have enough data to train our models, and have enough data to test them. The 80/20 split is preferred over the more common approach of 90/10, this is because our data set is very unbalanced and we need to have all different modes of failure in the test data.



## II.III. Machine Learning Model Algorithms
In this project we will use 7 different machine learning algorithms. The list below includes the Caret name of these algorithms, their actual names, and a short description of them:

**1) raprt (Decision tree)**: A machine learning model that is based on iterative partitioning of feature space and assigning labels to each region.\newline
**2) rf (Random forest)**: An ensemble of multiple decision trees. The output of the classification is the output chosen the most by the trees in the ensemble.\newline
**3) lda2 (Linear discriminant analysis)**: A machine learning algorithms that tries to linearly divide the data after improving the separability of classes by projecting the data into the lower dimensional linear discriminant (LD) space by maximizing the cross-class distances, while minimizing the in-class variation.\newline
**4) pda2 (Penalized discriminant analysis)**: an extension/variation of the LDA method, which incorporates penalty terms in the loss function for improving the generalizability of the model and prevention of overfitting.\newline
**5) nb (Naive Bayes)**: A machine learning algorithm based on Bayes' theorem. This method creates the a priori probabilities based on the original distribution of the classes in the training set, and updates those probabilities based on observation of other features.\newline
**6) xgbTree (Extreme gradient boosting)**: A greatly optimized version of gradient boosting algorithm, which is effectively an ensemble of weak learners (e.g. trees) trained in a sequential manner, where each learner improves on the error made by the ensemble thus far.\newline
**7) LogitBoost (Boosted logistic regression)**: A method that operates in a comparable way to gradient boosting, and uses an ensemble of logistic regression models which are trained sequentially and learn from mistakes of the ensemble.\newline

It is worth mention mentioning that we use 5 fold cross-validation with default parameters for tuning of all models during the training:

```{r, cross validation, echo=TRUE}
# 5 fold cross validation
ctrl <- trainControl(method = "cv", number = 5) 
```

## II.IV. Other Utilized Methods 
To improve the performance of the models through different iterations of modelling, some additional techniques are used int this project. Here, we list these methods and provide a short description for each one:

**1) Synthetic Minority Oversampling Technique (SMOTE)**: A technique to generate synthetic data for the minority class in unbalanced data sets by interpolating between the existing instances of the minority class [4].\newline
**2) Density Based Synthetic Minority Oversampling Technique (DBSMOTE)**: An extension of SMOTE where synthetic instances are created based on the density of the minority class in each region [2]. DBSMOTE tries to overcome some undesired artifact caused by SMOTE (e.g. building bridges of data points between instances of the minority class that are far away from the main cluster of minority class).\newline
**3) Edited Nearest Neighbor (ENN)**: An undersampling method aimed at reducing the noise in the data by removing the instances which are surrounded by the "neighbors" from other classes [3]. \newline
**4) Projection to LD space**: Projecting the feature space into a lower dimensional linear discriminant (LD) space to improve the separability of the classes and minimize the in-class variations.

## II.V. Use of Functions in R
In order to have a cleaner script and to avoid copy-pasting big chunks of code, we have defined many useful functions in the R script of this project. Through out this report, you may encounter some of these functions. The interested reader is encouraged to inspect these functions in the R script provided alongside with this report.

\newpage

# III. Results
In this section, we are going to go through the details and results of different approaches and models during different iterations of model development. Firstly, we are going to start with a simple binary classification approach and compare the performance of different algorithms in the simple case. Then we extend our efforts to multi-class classification, and we also employ some techniques to mitigate the effect of the unbalance and the noise in the data. Later, we try an approach to improve the separability of classes by means of transformations and projections, as well as engineering of physically more meaningful features. Finally, we create an ensemble of our the best performing models. The metrics described in the introduction section are used to evaluate model performance and guide our efforts through different iterations of model development.

## III.I. Binray Classification
In the first iteration, we treat the  problem as an entirely binary classification problem, so we will use the **failure** columns as our target. Additionally, we will use all the relevant features present in the data set:
```{r target and feature definition binary classifier, echo=TRUE}
# features to use for training and target to predict
my_features <- c("air_t", "process_t", "rpm", "torque", 
                 "wear", "type_H", "type_L", "type_M")

my_target   <- "failure"

```
The training will include all the 7 algorithms described in the analysis section, and will take place using "train_model" function defined in our R script:
```{r method definitions and binary trainings, echo=TRUE, warning=FALSE, message=FALSE}
# methods and respective names to train models
methods      <- c("rpart", "rf", "lda2","pda2", "nb", "xgbTree",  "LogitBoost")
mehtod_names <- c("Decision tree", "Random forest", "LDA", "Penalized DA",
                  "Naive Bayes", "XGBoost", "Boosted logistic regression")

# train the models in lapply
invisible(capture.output(
  trained_models_binary <- lapply(methods, function(method){
  
    model <- train_model(df_training=train_set, 
              target=my_target,
              features=my_features, 
              method=method,
              control=ctrl)

  model
  })
)) 
```
After the training, we can plot the ROC curves for the models to get an idea of the performance of each algorithm. This takes place using the "plot_ROC_curves" function defined in the R script, which uses test set to get the probabilities for each prediction, and extract the TPR and FPR before plotting the curves:
```{r ROC for binary classification, warning=FALSE, message=FALSE, echo=TRUE, fig.height=6, fig.width=10}
plot_ROC_curves(trained_models=trained_models_binary,
                model_names=mehtod_names, 
                df_test=test_set, 
                target=my_target,
                title= "ROC curves comparison for binary classification") 
```
It can be seen that random forest and XGBoost are performing better than other models. The table below shows the TPR at 10% FPR for each of the models:
```{r TPR for binray, echo=FALSE}
# get our metric
tpr_at_10fpr <- sapply(methods, function(method){
  
  tprs <- get_tpr_at_x_fpr(trained_models_binary[[which(methods==method)]],
                           df_test=test_set, 
                           target=my_target, x=0.1)
  round(tprs, 4)
  
})

# create the table
tpr_table <- matrix(unname(tpr_at_10fpr), nrow = 1, byrow = TRUE)
rownames(tpr_table) <- c("binary classification")
colnames(tpr_table) <- names(tpr_at_10fpr)
knitr::kable(tpr_table)
```
And here are the F1-scores:
```{r F1 for binray, echo=FALSE}
# get f1-scores
f1_scores <- sapply(methods, function(method){
  
  f1 <- get_f1_score(
    get_confusion_matrix(
      trained_models_binary[[which(methods==method)]],
      df_test=test_set,
      target="failure")$table)
  round(f1, 4)
  
})

# create the table
f1_table <- matrix(unname(f1_scores), nrow = 1, byrow = TRUE)
rownames(f1_table) <- c("binary classification")
colnames(f1_table) <- names(f1_scores)
knitr::kable(f1_table)
```
The F1-scores are not really in the desired range yet. Therefore, we need to take another approach.

## III.II. Multi-Class Classification
In this iteration, we will train the our models using the same ML algorithms, with the exception that this time we are going to train them to classify the **failure_type** columns instead of the **failure** column: 
```{r multicalss training, echo=TRUE}
# multi-class classification
my_multiclass_target   <- "failure_type"

# train the models in lapply
invisible(capture.output(
  trained_models_multiclass <- lapply(methods, function(method){
    print(method)
    model <- train_model(df_training=train_set, 
                         target=my_multiclass_target,
                         features=my_features, 
                         method=method,
                         control=ctrl)
    model
  })
))

```
For evaluating the model performance, we will not distinguish between different positive classes (i.e. if the model detected the failure but the mode of the failure is wrong, we still count it as a true positive). Consequently, the ROC curves look like this:

```{r ROC for multiclass, echo=FALSE, fig.height=6, fig.width=10}
my_binary_target   <- "failure"
plot_ROC_curves(trained_models=trained_models_multiclass,
                model_names=mehtod_names, 
                df_test=test_set, 
                target=my_binary_target, 
                title="ROC curves comparison for multi-class classification") 

```
The TPR metric at 10% FPR shows substantial improvement for some of the models:
```{r TPR multiclass, echo=FALSE}
##
# get new tpr scores at 10% fpr
new_tpr_row <- unname(sapply(methods, function(method){
    tpr <- get_tpr_at_x_fpr(trained_models_multiclass[[which(methods==method)]],
                       df_test=test_set, 
                       target=my_binary_target, x=0.1)
    round(tpr, 4)}))

# add them to the table
tpr_table <- rbind(tpr_table, new_tpr_row)    
rownames(tpr_table)[nrow(tpr_table)] <- "multiclass classification"
knitr::kable(tpr_table)

```
The F1-scores changed for each model in different ways. This shows that we need more improvements to get to the desired levels of model performance:
```{r F1 for multiclass, echo=FALSE}
##
# get new f1-scores
new_f1_row <- unname(
  sapply(methods, function(method){
  f1 <- get_f1_score(
    conver_to_binary_confusion_matrix(
    get_confusion_matrix(
      trained_models_multiclass[[which(methods==method)]],
      df_test=test_set,
      target=my_multiclass_target)$table))
  round(f1, 4)}))

# add them to the table
f1_table <- rbind(f1_table, new_f1_row)    
rownames(f1_table)[nrow(f1_table)] <- "multiclass classification"
knitr::kable(f1_table)
```
Here is the comparison of the ROC curves compared to the binary classification case:

```{r ROC comparison multicalss vs binray, echo=FALSE, fig.height=6, fig.width=10}
plot_ROC_curves_cross_comparison(trained_models_1 = trained_models_binary, 
                                 trained_models_2 = trained_models_multiclass,
                                 set_1_description = "binary",
                                 set_2_description = "multiclass",
                                 method_names = c("rf","xgbTree"),
                                 model_names = c("Random forest", "XGBoost"), 
                                 df_test=test_set, target=my_binary_target)
```

To find out the weak point of our models, we can look at the confusion matrices for the best performing models. The following tables are the confusion matrices for the XGBoost and the random forest models respectively:
```{r confusion matrix multiclass, echo=FALSE}
knitr::kable(
get_confusion_matrix(trained_models_multiclass[[which(methods=="xgbTree")]],
                     df_test=test_set,
                     target=my_multiclass_target)$table
)

knitr::kable(
get_confusion_matrix(trained_models_multiclass[[which(methods=="rf")]],
                     df_test=test_set,
                     target=my_multiclass_target)$table
)

```
It seems to be the case that the "Tool Wear Failure" is the most difficult type of failure to detect. This may be related to the fact that "Tool Wear Failure" is the least common type of failure in our data set, as shown in the analysis section. To overcome this problem, we will employ SMOTE as described previously in this report.

## III.III. Multi-Class Classification with SMOTE
For applying SMOTE on our data set, we will use the "SMOTE" function of the "smotefamily" package in R. However, since this function is designed for binary classifications, we define another function that applies SMOTE to each minority class separately. Although we will oversample the minorities, it is not recommended to oversample them to the point of having a completely balanced data set. This is because some of the models use the a priori probabilities in their predictions, and since we are oversampling only in training set to avoid data leakage, once these models are employed to predict on the test set, they may have too many false positives. After applying SMOTE to our training set, we can see how the distribution of the minority classes has changed:
```{r, smote parameters and apply smote, include=FALSE}
my_multiclass_target <- "failure_type"
my_binary_target <- "failure"
positive_class <- "Failure"
n_subclass_synth <- 500

train_set_subclass_smote <- smote_for_failure_type(
  train_data=train_set, 
  target_col=my_binary_target, 
  positive_class=positive_class,
  subclass_type_col=my_multiclass_target, 
  features_cols=my_features,
  apprx_subclass_pop=n_subclass_synth)
```
```{r showing distributions after smote, echo=TRUE}
# before SMOTE
knitr::kable(table(train_set$failure_type), 
             col.names = c("Failure Type","Count"))
# after SMOTE
knitr::kable(table(train_set_subclass_smote$failure_type), 
             col.names = c("Failure Type","Count"))
```
As it can be seen, the distribution has changed and now we have around 500 positive sample for each minority class. After we train our models on this modified data set, we get the following ROC curves:

```{r training multiclass SMOTE, echo=FALSE, include=FALSE}
# train the models based on new training set from SMOTE
trained_models_subclass_SMOTE <- lapply(methods, function(method){
  print(method)
  model <- train_model(df_training=train_set_subclass_smote, 
                       target=my_multiclass_target,
                       features=my_features, 
                       method=method,
                       control=ctrl)
  model
})
```
```{r ROC multiclass SMOTE, echo=FALSE, fig.height=6, fig.width=10}
# plot the new ROC curves after SMOTE
plot_ROC_curves(trained_models=trained_models_subclass_SMOTE,
                model_names=mehtod_names, 
                df_test=test_set, 
                target=my_binary_target,
                title="ROC curves comparison for multi-class classification after SMOTE") 
```
The TPR metric at 10% FPR after SMOTE is applied shows us that some models like decision tree and LDA show significant improvement:
```{r TPR for multiclass SMOTE, echo=FALSE}
# get new tpr scores at 10% fpr
new_tpr_row <- unname(sapply(methods, function(method){
  tpr <- get_tpr_at_x_fpr(trained_models_subclass_SMOTE[[which(methods==method)]],
                          df_test=test_set, 
                          target=my_binary_target, x=0.1)
  round(tpr, 4)}))

# add them to the table
tpr_table <- rbind(tpr_table, new_tpr_row)    
rownames(tpr_table)[nrow(tpr_table)] <- "multiclass + SMOTE"
knitr::kable(tpr_table)
```
Although the F1-scores after SMOTE seems to have improved for most models, but it did not improved a lot for random forest, and it even got worse for XGBoost:
```{r F1 scores multiclass SMOTE, echo=FALSE}
# get new f1-scores
new_f1_row <- unname(
  sapply(methods, function(method){
    f1 <- get_f1_score(
      conver_to_binary_confusion_matrix(
        get_confusion_matrix(
          trained_models_subclass_SMOTE[[which(methods==method)]],
          df_test=test_set,
          target=my_multiclass_target)$table))
    round(f1, 4)}))

# add them to the table
f1_table <- rbind(f1_table, new_f1_row)    
rownames(f1_table)[nrow(f1_table)] <- "multiclass + SMOTE"
knitr::kable(f1_table)
```
The following figure shows the ROC curves before and after applying SMOTE:

```{r ROC comparison smote vs multiclass, echo=FALSE, fig.height=6, fig.width=10}
plot_ROC_curves_cross_comparison(trained_models_1 = trained_models_multiclass, 
                                 trained_models_2 = trained_models_subclass_SMOTE,
                                 set_1_description = "multiclass",
                                 set_2_description = "multiclass SMOTE",
                                 method_names = c("rf","xgbTree"),
                                 model_names = c("Random forest", "XGBoost"), 
                                 df_test=test_set, target=my_binary_target)
```

We can have a look at the confusion matrices to see if SMOTE has helped in detecting the "Tool Wear Failure". The following tables are the confusion matrices for the XGBoost and the random forest models respectively:
```{r consusion matrix SMOTE, echo=FALSE}
# show confusion matrix for XGboost and random forest
knitr::kable(
get_confusion_matrix(trained_models_subclass_SMOTE[[which(methods=="xgbTree")]],
                     df_test=test_set,
                     target=my_multiclass_target)$table
)
knitr::kable(
get_confusion_matrix(trained_models_subclass_SMOTE[[which(methods=="rf")]],
                     df_test=test_set,
                     target=my_multiclass_target)$table
)
```
As it can be seen, we still fail in detecting the "Tool Wear Failure" instances. The application of SMOTE did not drastically changed the performance of our best models. Therefore, Further improvements are necessary.

## III.IV. Multi-Class Classification with DBSMOTE
DBSMOTE is an extension of the regular SMOTE, but it usually performs better due to avoiding oversampling in noisy areas. We will employ DBSOMTE in a similar way as we did for SMOTE, and inspect the distribution of failure types before and after:
```{r DBSMOTE parameters, include=FALSE}
# NEXT: do density based smote (DBSMOTE) which doesn't let bridges 
# between main minority cluster and a random minority point.
my_multiclass_target <- "failure_type"
my_binary_target <- "failure"
positive_class <- "Failure"
n_subclass_synth <- 500

# apply DBSMOTE on the train_set
train_set_subclass_DBSMOTE <- DBSMOTE_for_failure_type(
  train_data=train_set, 
  target_col=my_binary_target, 
  positive_class = positive_class,
  subclass_type_col=my_multiclass_target, 
  features_cols=my_features,
  apprx_subclass_pop=n_subclass_synth)
```
```{r show distribution after DBSMOTE, echo=TRUE}
# distribution before DBSMOTE
knitr::kable(table(train_set$failure_type), 
             col.names = c("Failure Type","Count"))
# distribution after DBSMOTE
knitr::kable(table(train_set_subclass_DBSMOTE$failure_type), 
             col.names = c("Failure Type","Count"))
```
Using this new data set, we train our models and look at the new ROC curves:
```{r DBSMOTE training, include=FALSE}
# train the models based on new training set from DBSMOTE
trained_models_subclass_DBSMOTE <- lapply(methods, function(method){
  model <- train_model(df_training=train_set_subclass_DBSMOTE, 
                       target=my_multiclass_target,
                       features=my_features, 
                       method=method,
                       control=ctrl)
  model
})
```
```{r DBSMOTE ROC curves, echo=FALSE, fig.height=6, fig.width=10}
# plot the new ROC curves after DBSMOTE
plot_ROC_curves(trained_models=trained_models_subclass_DBSMOTE,
                model_names=mehtod_names, 
                df_test=test_set, 
                target=my_binary_target,
                title="ROC curves comparison for multi-class classification after DBSMOTE") 
```
The curves do not look very different from the last time. We can confirm this by checking out the TPR at 10% FPR for each model:
```{r TPR DBSMOTE, echo=FALSE}
# get new tpr scores at 10% fpr
new_tpr_row <- unname(sapply(methods, function(method){
  tpr <- get_tpr_at_x_fpr(trained_models_subclass_DBSMOTE[[which(methods==method)]],
                          df_test=test_set, 
                          target=my_binary_target, x=0.1)
  round(tpr, 4)}))

# add them to the table
tpr_table <- rbind(tpr_table, new_tpr_row)    
rownames(tpr_table)[nrow(tpr_table)] <- "multiclass + DBSMOTE"
knitr::kable(tpr_table)
```
Our F1-scores seem to have gotten worse in all cases except for the decision tree classifier:
```{r F1 scores DBSMOTE, echo=FALSE}
# get new f1-scores
new_f1_row <- unname(
  sapply(methods, function(method){
    f1 <- get_f1_score(
      conver_to_binary_confusion_matrix(
        get_confusion_matrix(
          trained_models_subclass_DBSMOTE[[which(methods==method)]],
          df_test=test_set,
          target=my_multiclass_target)$table))
    round(f1, 4)}))

# add them to the table
f1_table <- rbind(f1_table, new_f1_row)    
rownames(f1_table)[nrow(f1_table)] <- "multiclass + DBSMOTE"
knitr::kable(f1_table)
```
We can also compare the ROC curves before and after applying DBSMOTE:

```{r ROC comparison DMSMOTE vs multiclass, echo=FALSE, fig.height=6, fig.width=10}
plot_ROC_curves_cross_comparison(trained_models_1 = trained_models_multiclass, 
                                 trained_models_2 = trained_models_subclass_DBSMOTE,
                                 set_1_description = "multiclass",
                                 set_2_description = "multiclass DBSMOTE",
                                 method_names = c("rf","xgbTree"),
                                 model_names = c("Random forest", "XGBoost"), 
                                 df_test=test_set, target=my_binary_target)
```

Confusion matrices for XGBoost and random forest after applying DBSMOTE are shown below. Yet again, we failed to adequately detect the "Tool Wear Failure" instances. Therefore, we will take more measures to improve the model performance.
```{r confusion matrices DBSOMTE, echo=FALSE}
# show confusion matrix for XGboost and random forest
knitr::kable(
get_confusion_matrix(trained_models_subclass_DBSMOTE[[which(methods=="xgbTree")]],
                     df_test=test_set,
                     target=my_multiclass_target)$table)
knitr::kable(
get_confusion_matrix(trained_models_subclass_DBSMOTE[[which(methods=="rf")]],
                     df_test=test_set,
                     target=my_multiclass_target)$table)
```

## III.V. Multi-Class Classification with DBSMOTE/ENN
ENN, as described previously in this report, is used to de-noise the data by undersampling the instances in noisy regions in our feature space. Combination ENN with one of the variations of SMOTE is a pretty common practice in data science. Here we combine DBSMOTE and ENN to see if we can improve our model performance:
```{r ENN params, include=FALSE}
# Applying ENN on DBSMOTE database before training to reduce noise
my_multiclass_target <- "failure_type"
my_binary_target <- "failure"
positive_class <- "Failure"
```
```{r ENN apply, echo=TRUE}
# apply ENN
train_set_subclass_DBSMOTE_ENN <- apply_ENN(df=train_set_subclass_DBSMOTE,
                                            target=my_multiclass_target,
                                            features=my_features)
```
The distribution of failure type after applying ENN to the data set produced by DBSOMTE, we can see that some instances have been removed from the data set as expected.
```{r ENN distributions, echo=TRUE}
# distribution before DBSMOTE/ENN
knitr::kable(table(train_set_subclass_DBSMOTE$failure_type), 
             col.names = c("Failure Type","Count"))
# distribution after DBSMOTE/ENN
knitr::kable(table(train_set_subclass_DBSMOTE_ENN$failure_type), 
             col.names = c("Failure Type","Count"))
```
After we using this modified data set to train our models, we get the following ROC curves from our trained models:
```{r ENN model taining, include=FALSE}
# train models
trained_models_subclass_DBSMOTE_ENN <- lapply(methods, function(method){
  print(method)
  model <- train_model(df_training=train_set_subclass_DBSMOTE_ENN, 
                       target=my_multiclass_target,
                       features=my_features, 
                       method=method,
                       control=ctrl)
  model
})
```
```{r, ENN ROC curves, echo=FALSE, echo=FALSE, fig.height=6, fig.width=10}
# plot the new ROC curves after DBSMOTE_ENN
plot_ROC_curves(trained_models=trained_models_subclass_DBSMOTE_ENN,
                model_names=mehtod_names, 
                df_test=test_set, 
                target=my_binary_target,
                title= "ROC curves comparison for multi-class classification after DBSMOTE + ENN") 
```
We can look at our metrics to see if the model performance has improved. The TPR values at 10% FPR show us that ENN did not improve model performances;
```{r TPR ENN, echo=FALSE}
# get new tpr scores at 10% fpr
new_tpr_row <- unname(sapply(methods, function(method){
  tpr <- get_tpr_at_x_fpr(trained_models_subclass_DBSMOTE_ENN[[which(methods==method)]],
                          df_test=test_set, 
                          target=my_binary_target, x=0.1)
  round(tpr, 4)}))

# add them to the table
tpr_table <- rbind(tpr_table, new_tpr_row)    
rownames(tpr_table)[nrow(tpr_table)] <- "multiclass + DBSMOTE/ENN"
knitr::kable(tpr_table)
```
The following table compares contains the updated F1-scores after DBSOMTE/ENN. We can clearly see the decrease in performance for our best 2 models. Therefore, we will stop using this strategy (DBSMOTE+ENN).
```{r F1 ENN, echo=FALSE}
# get new f1-scores
new_f1_row <- unname(
  sapply(methods, function(method){
    f1 <- get_f1_score(
      conver_to_binary_confusion_matrix(
        get_confusion_matrix(
          trained_models_subclass_DBSMOTE_ENN[[which(methods==method)]],
          df_test=test_set,
          target=my_multiclass_target)$table))
    round(f1, 4)}))

# add them to the table
f1_table <- rbind(f1_table, new_f1_row)    
rownames(f1_table)[nrow(f1_table)] <- "multiclass + DBSMOTE/ENN"
knitr::kable(f1_table)
```
The following plot compares the ROC curves before and after DBSMOTE/ENN for our 2 best models. We can see that the models perform worse after applying DBSMOTE/ENN to the data set.
```{r ENN ROC comparison, echo=FALSE, fig.height=6, fig.width=10}
# compare with original
plot_ROC_curves_cross_comparison(trained_models_1 = trained_models_multiclass, 
                                 trained_models_2 = trained_models_subclass_DBSMOTE_ENN,
                                 set_1_description = "multiclass",
                                 set_2_description = "multiclass DBSMOTE+ENN",
                                 method_names = c("rf","xgbTree"),
                                 model_names = c("Random forest", "XGBoost"), 
                                 df_test=test_set, target=my_binary_target)
```
Having a look at our confusion matrices for XGBoost and random forest, we can see that we did not succeed in detecting "Tool Wear Failure", and we have significantly more false positives:
```{r ENN confusion matrix, echo=FALSE}
# show confusion matrix for XGboost and random forest
knitr::kable(get_confusion_matrix(trained_models_subclass_DBSMOTE_ENN[[which(methods=="xgbTree")]],
                     df_test=test_set,
                     target=my_multiclass_target)$table)

knitr::kable(get_confusion_matrix(trained_models_subclass_DBSMOTE_ENN[[which(methods=="rf")]],
                     df_test=test_set,
                     target=my_multiclass_target)$table)
```

## III.VI. Multi-Class Classification in LD Space
As we have failed to improve the model performance by means of SMOTE, DBSMOTE, and ENN, it makes sense to try a different approach. As described in the analysis section of this report, we can project our features into a lower dimensional space of linear discriminants (LDs) to improve separability of our classes and minimize the in-class variations. The LD space is in fact the space where the LDA model classifies the instances. Here we try to classify instances in the LD space but using models other than LDA. First we project the training set onto the LD space:

```{r LD projection, echo=TRUE}
# extract features and target
train_set_pre_LDA <- train_set[c(my_features,my_multiclass_target)]  

# apply LDA
lda_result <- MASS::lda(failure_type ~ ., data = train_set_pre_LDA)

# put labels back in the dataframe
train_set_LDA <- data.frame(predict(lda_result, train_set_pre_LDA)$x)
train_set_LDA[[my_binary_target]] <- train_set[[my_binary_target]]
train_set_LDA[[my_multiclass_target]] <- train_set[[my_multiclass_target]]
```
After this projection, we project the test set using the same coefficients we obtained during the projection of the training set. This is done to avoid data leakage:
```{r LD projection test set, echo=TRUE}
# transform the test set with the same coefficient
test_set_LDA <- data.frame(predict(lda_result, test_set)$x)
test_set_LDA[[my_binary_target]] <- test_set[[my_binary_target]]
test_set_LDA[[my_multiclass_target]] <- test_set[[my_multiclass_target]]
```
Here is a plot of our training set data projected onto the plane defined by the first 2 linear discriminants:

```{r LD1 LD2 data plot, echo=FALSE, fig.height=4, fig.width=6}
test_set_LDA %>% ggplot(aes(LD1, LD2, color=failure)) + geom_point(alpha=0.5)
```

It seems like the data is more separable in LD space when we compare this plot to the ones provided in the analysis section.
Here are the models we are going to train on the data projected into the 4 dimensional LD space:
```{r LD space models, echo=TRUE}
# redefine methods
new_methods      <- c("rf", "xgbTree",  "LogitBoost")
new_mehtod_names <- c("Random forest","XGBoost", "Boosted logistic regression")
```
And after training our models, we obtain the following ROC curves:
```{r LD training models, include=FALSE}
# features to use for training
my_features_LDA <- c("LD1", "LD2", "LD3", "LD4")

trained_models_subclass_LDA <- lapply(new_methods, function(method){
  print(method)
  model <- train_model(df_training=train_set_LDA, 
                       target=my_multiclass_target,
                       features=my_features_LDA, 
                       method=method,
                       control=ctrl)
  model
})

```
```{r ROC LD space, echo=FALSE, fig.height=6, fig.width=10}
# plot the new ROC curves 
plot_ROC_curves(trained_models=trained_models_subclass_LDA,
                model_names=new_mehtod_names, 
                df_test=test_set_LDA, 
                target=my_binary_target,
                title= "ROC curves comparison for multi-class classification in LD space") 

```

It is not clear if we have an improvement. We can find out about that by looking into our metrics. The table below shows the TPR at 10% FPR:
```{r LD TPR, echo=FALSE}
# get new tpr scores at 10% fpr
new_tpr_row <- unname(sapply(seq_along(new_methods), function(i){

  tpr <- get_tpr_at_x_fpr(trained_models_subclass_LDA[[i]],
                          df_test=test_set_LDA, 
                          target=my_binary_target, x=0.1)
  round(tpr, 4)}))

# NA for the missing methods
new_tpr_row_modified <- rep(NA, length(methods))
new_tpr_row_modified[methods %in% new_methods] <- new_tpr_row

# add them to the table
tpr_table <- rbind(tpr_table, new_tpr_row_modified)    
rownames(tpr_table)[nrow(tpr_table)] <- "multiclass in LD space"
knitr::kable(tpr_table)
```
The F1-score values obtained from this approach show significant worsening of the performance of our models:
```{r LD F1, echo=FALSE}
# get new f1-scores
new_f1_row <- unname(
  sapply(seq_along(new_methods), function(i){
    f1 <- get_f1_score(
      conver_to_binary_confusion_matrix(
        get_confusion_matrix(
          trained_models_subclass_DBSMOTE_ENN[[i]],
          df_test=test_set,
          target=my_multiclass_target)$table))
    round(f1, 4)}
    )
  )

# NA for the missing methods
new_f1_row_modified <- rep(NA, length(methods))
new_f1_row_modified[methods %in% new_methods] <- new_f1_row

# add them to the table
f1_table <- rbind(f1_table, new_f1_row_modified)    
rownames(f1_table)[nrow(f1_table)] <- "multiclass in LD space"
knitr::kable(f1_table)
```

## III.VII. Multi-Class Classification + Feature Engineering
So far we have used different ML algorithms, oversampling techniques, de-noising the data, and projecting features onto lower transformed spaces. However, in every case we used the same set of features. Maybe the key to improve the model performance is to come up with better features. Coming from an engineering background, it was not difficult to define the following new features: \newline
**1) power:** Multiplication of rpm and toque. \newline
**2) strain:** The cumulative stain on the machine, calculated by multiplying the torque and the wear (in minutes). \newline
**3) delta_T:** The difference between process temperature and the air temperature.

Let us create the necessary columns in the training set and the test set:
```{r physical features, echo=TRUE}
# define new features which make physical sense
train_set <- train_set %>%
  mutate(strain=wear*torque,
         delta_T= process_t-air_t, 
         power=rpm*torque)

# define the same features for the test set
test_set <- test_set %>%
  mutate(strain=wear*torque, 
         delta_T= process_t-air_t, 
         power=rpm*torque)
```
We will use the same 7 ML algorithms, but we will change the features we will use for the training:
```{r methods and features for phisically meaningful part, echo=TRUE}
# methods and respective names to train models
methods      <- c("rpart", "rf", "lda2","pda2", "nb", "xgbTree",  "LogitBoost")
mehtod_names <- c("Decision tree", "Random forest", "LDA", "Penalized DA",
                  "Naive Bayes", "XGBoost", "Boosted logistic regression")

# features to use for training and target to predict
my_features <- c("delta_T", "power", "strain", "rpm", "torque", 
                 "wear", "type_H", "type_L", "type_M")

my_multiclass_target   <- "failure_type"
```
After training our models using the new features, we obtain the following ROC curves:
```{r training feature engineering, include=FALSE}
# train the models in lapply
trained_models_subclass_physical_features <- lapply(methods, function(method){
  print(method)
  model <- train_model(df_training=train_set, 
                       target=my_multiclass_target,
                       features=my_features, 
                       method=method,
                       control=ctrl)
  model
})
```
```{r ROC physical, echo=FALSE, fig.height=6, fig.width=10}
# plot ROC curves
my_binary_target <- "failure"
plot_ROC_curves(trained_models=trained_models_subclass_physical_features,
                model_names=mehtod_names, 
                df_test=test_set, 
                target=my_binary_target,
                title="ROC curves comparison for multi-class classification using the new features") 
```
We can already see a significant improvement in our ROC curves. We can confirm the improvement in our models by calculating the TPR at 10% FPR for each model
```{r TPR feature engineering, echo=FALSE}
# get new tpr scores at 10% fpr
new_tpr_row <- unname(sapply(methods, function(method){
  tpr <- get_tpr_at_x_fpr(trained_models_subclass_physical_features[[which(methods==method)]],
                          df_test=test_set, 
                          target=my_binary_target, x=0.1)
  round(tpr, 4)}))
# add them to the table
tpr_table <- rbind(tpr_table, new_tpr_row)    
rownames(tpr_table)[nrow(tpr_table)] <- "multiclass + new features"
knitr::kable(tpr_table)
```
We see significant improvement for most of the models. We can see similar results in our F1-scores:
```{r F1 feature engineering, echo=FALSE}
# get new f1-scores
new_f1_row <- unname(
  sapply(methods, function(method){
    f1 <- get_f1_score(
      conver_to_binary_confusion_matrix(
        get_confusion_matrix(
          trained_models_subclass_physical_features[[which(methods==method)]],
          df_test=test_set,
          target=my_multiclass_target)$table))
    round(f1, 4)}))
# add them to the table
f1_table <- rbind(f1_table, new_f1_row)    
rownames(f1_table)[nrow(f1_table)] <- "multiclass + new features"
knitr::kable(f1_table)
```
The F1-scores are now very close to the desired level mentioned in the introduction section. We can also visually compare the model performances before and after adding the new features by looking at the ROC curves:
```{r ROC comparison feature engineering, echo=FALSE, fig.height=6, fig.width=10}
# compare the ROC curves of new approach to the ones of original multiclass models
plot_ROC_curves_cross_comparison(trained_models_1 = trained_models_multiclass, 
                                 trained_models_2 = trained_models_subclass_physical_features,
                                 set_1_description = "multiclass",
                                 set_2_description = "multiclass new features",
                                 method_names = c("rf","xgbTree"),
                                 model_names = c("Random forest", "XGBoost"), 
                                 df_test=test_set,
                                 target=my_binary_target)
```
And here are the confusion matrices for XGBoost and random forest after adding the new features to our data set:
```{r confusion matrix new features, echo=FALSE}
# show confusion matrix for XGboost and random forest
knitr::kable(get_confusion_matrix(trained_models_subclass_physical_features[[which(methods=="xgbTree")]],
                     df_test=test_set,
                     target=my_multiclass_target)$table)

knitr::kable(get_confusion_matrix(trained_models_subclass_physical_features[[which(methods=="rf")]],
                     df_test=test_set,
                     target=my_multiclass_target)$table)
```
Although we still cannot address the "Tool Wear Failure" class as adequately as we may desire, but the rest of the cases seem pretty good in general. At this point, we can stop trying new techniques, and make an ensemble of our best models.

## III.VIII. The Ensemble
We will use our 3 best models from the last section to create the ensemble, and we will use hard voting to carry out the classification task. The multi-class confusion matrix of the ensemble model is provided below:
```{r ensemble predict, echo=TRUE}
# models to be used in ensemble
ensemble_methods <- c("rf","xgbTree", "LogitBoost")

# predict based on voting of models in ensemble
ensemble_predictions <- predict_ensemble(
  models=trained_models_subclass_physical_features, 
  ensemble_methods=ensemble_methods,
  new_data=test_set)

# assemble confusion matrix
ensemble_prediction_table <- table(ensemble_predictions, test_set$failure_type)
knitr::kable(ensemble_prediction_table)
```
We can see that in total we have 13 false negatives, all belonging to the "Tool Wear Failure" class. And we have no False positives. Here is the binary confusion matrix of the ensemble model:
```{r binary confusion matrix ensemble, echo=FALSE}
# convert multi-class to binary confusion matrix
binary_confusion_matrix <- conver_to_binary_confusion_matrix(ensemble_prediction_table)
knitr::kable(binary_confusion_matrix)
```
We can easily calculate our final F1-score:
```{r F1 score}
# calculate f1-score
get_f1_score(binary_confusion_matrix)
```
This is very close to our stated goal in the introduction section, which was a F1-score of 0.9. Therefore, further iterations of the model is not needed. The plot below shows the maximum F1-scores in each iteration of model development, with the last point being the ensemble model:

```{r F1 evolution, echo=FALSE, fig.height=6, fig.width=10}
max_F1s <- apply(f1_table, 1, max, na.rm = TRUE)
max_F1s <- unname(max_F1s)
max_F1s <- c(max_F1s, get_f1_score(binary_confusion_matrix))

plot(0, 0, type = "n", xlim = c(1, 8), ylim = c(0, 1), 
      xlab = "Iteration", ylab = "F1-score", 
      main = "Evolution of F1-scores for the best performing model in each iteration")
grid(col = "grey", lty = "dotted")
lines(max_F1s, col = "blue3", lwd = 2)
```

Since the final model is an ensemble of 3 models, it would be a cumbersome operation to derive an ROC curve for it. For the same reason, we do not state a TPR at 10% FPR for the final model. 

\newpage

# IV. Conclusions
In this project, we developed the basis for a predictive maintenance model for industrial machines given in our data set. Firstly, we investigated the data set, observed the unbalance within the target columns, and did some data cleaning. Then, we carried out an exploratory data analysis (EDA) and made the necessary visualizations on the data set to find out about correlations within our data set. Then we started training models using 7 different ML algorithms. We applied different techniques such as SMOTE, DBSMOTE, ENN, projection onto LD space, and feature engineering in order to enhance our model performance and the F1-score. Once we brought the F1-score to the desired level, we created an ensemble of our best models and presented the results for the final model. The final model demonstrates perfect prediction capabilities on the test set with the exception the of "Tool Wear Failure" class. To improve this, one can try to train better models by defining the proper grid search parameters and tuning the hyper parameters of the model. Another possible way to improve the model would be to create a digital twin simulation model of the monitored system based on physical principles and use it to generate synthetic data for the failure class which is difficult to detect. This approach will probably result in better synthetic samples than the ones generated by SMOTE or DBSMOTE. 

\newpage

# V. References

[1] https://www.kaggle.com/datasets/shivamb/machine-predictive-maintenance-classification (accessed on 15.02.2024, at 20:37 CET)

[2] Bunkhumpornpat, C., Sinapiromsaran, K. & Lursinsap, C. DBSMOTE: "Density-Based Synthetic Minority Over-sampling TEchnique." Appl Intell 36, 664684 (2012). https://doi.org/10.1007/s10489-011-0287-y

[3] Wilson, Dennis L.. Asymptotic Properties of Nearest Neighbor Rules Using Edited Data. IEEE Trans. Syst. Man Cybern. 2 (1972): 408-421.

[4] N. V. Chawla, K. W. Bowyer, L. O. Hall, W. P. Kegelmeyer "SMOTE: Synthetic Minority Over-sampling Technique." Journal Of Artificial Intelligence Research, Volume 16, pages 321-357, 2002

